Machine Learning Methods for Discourse Segmentation of Communications in E-Mail Based Behavioral Interventions

### 2018-07-19 paper-amia-summit-2018/manuscript.tex
Added files for initial version of amia summit paper

\noindent{\bf Abstract}
\textit{Communication science approaches to developing effective behavior interventions, such as motivational interviewing (MI), are limited by traditionally manual qualitative coding of communication exchanges, which is a very resource-intensive and time-consuming process. This study focuses on the analysis of e-Coaching sessions, behavior interventions that are delivered via email and grounded in the principles of MI. A critical step towards automated annotation of e-Coaching communication exchanges is segmentation of emails into textual fragments that correspond to MI behaviors. In this work, we formulate this task as a classification problem and propose lexical, punctuation and topic features to address it. We experimented both with traditional machine learning methods, such as Support Vector Machine (SVM), Naive Bayes and K-Nearest Neighbor (KNN) classifiers and recurrent neural networks (RNNs). Results indicate that SVM outperformed KNN and RNNs achieving 0.990 macro F1-score overall and 0.848 and 0.995 macro F1-score for detecting ``new segment'' and ``same segment'' classes, respectively.}

@@

\textbf{Support Vector Machine}: state-of-the-art supervised classification method proven to be effective for text categorization\cite{joachims1998text} for its ability to cope with high dimensional input feature space. SVM finds the hyperplane that maximizes the separation between the closest ``new segment'' and ``same segment'' training examples. We used L1-regularized linear SVM with squared error minimization in experiments.

@@

\begin{table}[ht]
\centering
\caption{Performance of NB, SVM and KNN methods for identification of ``new segment'' and ``same segment'' classes as well as their weighted average. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_base}
  \begin{tabular}{|l|l|l|l|l|l|l|l|l|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{new segment}} & \multicolumn{3}{|c|}{\textbf{same segment}}  & \multicolumn{3}{|c|}{\textbf{overall}} \\\cline{2-10}
   & \textbf{Precision}  & \textbf{Recall} & \textbf{F1} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1}\\ \hline    
 \multicolumn{10}{|c|}{lexical features only} \\ \hline   
 NB & 0.594 & 0.662 & 0.626 & 0.988 & 0.985 & 0.987 & 0.975 & 0.974 & 0.975  \\ \hline
 SVM & 0.762 & \textbf{0.673} & 0.715 & \textbf{0.989} & 0.993 & 0.991  & 0.981 & 0.982 & 0.982 \\ \hline
 KNN & \textbf{0.808} & 0.663 & \textbf{0.728} & \textbf{0.989} & \textbf{0.995} & \textbf{0.992} & \textbf{0.983} & \textbf{0.984} & \textbf{0.983} \\ \hline
 \multicolumn{10}{|c|}{lexical and puctuation features} \\ \hline
 LSTM & \textbf{0.810} & 0.720 & \textbf{0.762}  & \textbf{0.994} & \textbf{0.994} & \textbf{0.994} & \textbf{0.988} & \textbf{0.985} & \textbf{0.987} \\ \hline
 GRU & 0.783 & \textbf{0.739} & 0.760 & \textbf{0.994} & \textbf{0.994} & \textbf{0.994} & \textbf{0.988} & \textbf{0.985} & \textbf{0.987} \\ \hline
  \end{tabular}
\end{table}

As follows from Table~\ref{tab:result_base}, KNN outperforms all other models in terms of precision and F1-measure achieving 0.808 precision with 0.728 F1-measure for new segment detection. KNN also shows superior performance in all performance metrics for ``same segment'' class and weighted average over ``new segment'' and ``same segment'' classes. NB demonstrates the lowest performance among all models in terms of all performance metrics. On the other hand, SVM has the highest recall of 0.673 when only lexical features are used to identify ``new segment''. Results indicate that performance of classifiers is remarkably higher for ``same segment'' class compared to ``new segment'' class, which is expected since 96.71\% instances belong to ``same segment'' categories. For example, KNN achieves 22.40\%, 50.07\% and 36.26\% higher precision, recall and F1-measure, respectively, in ``same segment'' identification compared to ``new segment'' detection. When lexical features are used in combination with punctuation mark, LSTM demonstrates the best performance in terms of precision and F1-measure for ``new segment'' identification while GRU shows 2.64\% higher recall than LSTM. GRU and LSTM show exactly same performance in terms of all performance metrics for ``same segement'' and overall classifications. \\

@@

 NB & 0.573 & 0.680 & 0.622 & 0.573 & 0.677 & 0.621 & 0.593 & 0.662 & 0.626 \\ \hline
 SVM & 0.800 & 0.799 & 0.799 & \textbf{0.848} & \textbf{0.849} & \textbf{0.848} & 0.836 & 0.839 & 0.837  \\ \hline
 KNN & 0.808 & 0.741 & 0.773 & 0.814 & 0.732 & 0.771 & 0.814 & 0.740 & 0.775 \\ \hline
  \end{tabular}
\end{table}

Table~\ref{tab:result_boundary} summarizes the performance of NB, SVM and KNN models for detecting ``new segment'' class. We observed that classification performance significantly improved for ``new segment'' class when lexical features are used in combination with punctuation and model-based topic features. Precision increases by 4.99\% and 0\%, 11.29\% and 0.74\% and 9.71\% and 0.74\% for SVM and KNN; recall increases by 18.72\% and 17.06\%, 26.15\% and 15.64\% and 24.67\% and 16.90\% for SVM and KNN; and F1-measure increases by 11.75\% and 6.18\%, 18.60\% and 5.91\% and 17.06\% and 6.46\% for SVM and KNN when L-LDA, DL-LDA and LCA-based topic features are used in conjunction combination with lexical and punctuation features, respectively. On the other hand, NB has very negligible impact of punctuation and model-based topic features. In contrast to results in Table~\ref{tab:result_base}, SVM demonstrates the best performance among all models achieving 0.848, 0.849 and 0.848 for precision, recall and F1-measure, respectively, when DL-LDA-based topic features are used in conjunction with lexical and punctuation features. NB and KNN provide their best performance when LCA-based topic features are used, achieving 0.626 and 0.775 F1-measure, respectively.\\

@@

 NB & 0.989 & 0.983 & 0.986 & 0.989 & 0.983 & 0.986 & 0.988 & 0.985 & 0.987 \\ \hline
 SVM & 0.993 & 0.993 & 0.993 & \textbf{0.995} & \textbf{0.995} & \textbf{0.995} & \textbf{0.995} & 0.994 & 0.994  \\ \hline
 KNN & 0.991 & 0.994 & 0.993 & 0.991 & 0.994 & 0.993 & 0.991 & 0.994 & 0.993 \\ \hline
  \end{tabular}
\end{table}

Similar to baseline performance, classification performance is remarkably higher for ``same segment'' class compared to ``new segment'' class (Table~\ref{tab:result_not_boundary}). The impact of punctuation and topic model-based features is also consistent with ``new segment'' classification while improvements are not as significant as ``new segment'' class. We observed that precision increases by 0.4\% and 0.2\%, 0.6\% and 0.2\% and 0.6\% and 0.2\% for SVM and KNN; recall increases by 0\% and -0.1\%, 0.2\% and -0.1\% and 0.1\% and -0.1\% for SVM and KNN; and F1-measure increases by 0.2\% and 0.1\%, 0.4\% and 0.1\% and 0.3\% and 0.1\% for SVM and KNN when L-LDA, DL-LDA and LCA-based topic features are used in conjunction combination with lexical and punctuation features, respectively. Similar to results in Table \ref{tab:result_boundary}, NB and KNN demonstrate the lowest and moderate performance on the task of identifying ``same segment'', respectively. \\

@@

 NB & 0.975 & 0.973 & 0.974 & 0.975 & 0.973 & 0.974 & 0.975 & 0.974 & 0.975 \\ \hline
 SVM & 0.987 & 0.987 & 0.987 & \textbf{0.990} & \textbf{0.990} & \textbf{0.990} & 0.989 & 0.989 & 0.989 \\ \hline
 KNN & 0.985 & 0.986 & 0.985 & 0.985 & 0.986 & 0.985 & 0.985 & 0.986 & 0.986 \\ \hline
  \end{tabular}
\end{table}

Table~\ref{tab:result_weighted_avg} summarizes the weighted average results of the models for segmentation of e-Coaching emails. Overall, SVM has the best performance across all topic model-based features in terms of all performance metrics, achieves 0.990 precision, recall and F1-measure when DL-LDA-based topic features are used. Similar to results in Tables~\ref{tab:result_base}, \ref{tab:result_boundary} and \ref{tab:result_not_boundary}, NB demonstrates the lowest performance among all methods while shows best performance with LCA-based topic features compared to other model-based topic features. Influence of the additional features is also consistent with the results in Tables~\ref{tab:result_boundary} and~\ref{tab:result_not_boundary}. Precision increases by 0.9\% and 0.2\%; recall increases by 0.8\% and 0.2\%; and F1-measure increases by 0.8\% and 0.3\% for SVM and KNN methods, respectively, when lexical feature is used in combination with punctuation and best model-based topic features.\\

\section*{Discussion}
This study is the first effort to evaluate the automatic segmentation of e-Coaching emails. Experimental results indicate that SVM is the best model among all machine learning methods considered for this study. SVM achieved 0.990 F1-measure in overall, 0.848 and 0.995 F1-measures for detecting ``new segment'' and ``same segment'', respectively. The robust performance of SVM provides the evidence that machine learning models are capable to learn conceptual information from clinical exchange. It also indicates that topic features are more important than lexical features, even when deep learning methods are employed. Although the domain of this study was intentionally quite small, we believe that our study is not limited to the e-Coaching domain and it can be successfully applied to other domains, in which discourse segmentation is a preliminary step for annotation.

@@

In this paper, results are reported by each class to avoid confusion regarding overall model performance due to severe class imbalance. In addition, standard metrics: precision, recall and F1-measure were used to eliminate doubt about the model performance since accuracy is misleading for imbalanced datasets. AUC is also utilized due to its effectiveness in measuring the quality of binary classifiers for imbalanced datasets\cite{hu2015kernelized}, which are demonstrated by the ROC curves in Figure~\ref{fig:roc-curves}. Although NB provides lowest classification results, it shows the second highest AUC values, achieving 0.978 AUC when only lexical features are used and 0.977 AUC when a combination of lexical, punctuation and topic features are used. On the other hand, KNN and SVM exhibit 0.972 and 0.833 AUCs when only lexical features are used; and 0.966 and 0.922 AUCs when a combination of lexical, punctuation and topic model-based features are used. Finally, LSTM demonstrates the highest AUC value among all machine learning models, achieving 0.996 AUC, while GRU achieves only 0.2\% lower AUC than LSTM.

>>>>

\noindent{\bf Abstract}
\textit{Communication science approaches to developing effective behavior interventions, such as motivational interviewing (MI), are limited by traditionally manual qualitative coding of communication exchanges, which is a very resource-intensive and time-consuming process. This study focuses on the analysis of e-Coaching sessions, behavior interventions that are delivered via email and grounded in the principles of MI. A critical step towards automated annotation of e-Coaching communication exchanges is segmentation of emails into textual fragments that correspond to MI behaviors. In this work, we formulate this task as a classification problem and propose lexical, punctuation and topic features to address it. We experimented both with traditional machine learning methods, such as Support Vector Machine (SVM), Naive Bayes and K-Nearest Neighbor (KNN) classifiers and recurrent neural networks (RNNs). Results indicate that SVM outperformed KNN and RNNs achieving 0.990 macro F1-score overall and 0.845 and 0.995 macro F1-score for detecting ``new segment'' and ``same segment'' classes, respectively.}

@@

\textbf{Support Vector Machine}: state-of-the-art supervised classification method proven to be effective for text categorization\cite{joachims1998text} for its ability to cope with high dimensional input feature space. SVM finds the hyperplane that maximizes the separation between the closest ``new segment'' and ``same segment'' training examples. We scaled our dataset to utilize linear SVM with optimal cost 512 in experiments.

@@

\begin{table}[ht]
\centering
\caption{Performance of NB, SVM and KNN methods for identification of ``new segment'' and ``same segment'' classes as well as their weighted average. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_base}
  \begin{tabular}{|l|l|l|l|l|l|l|l|l|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{new segment}} & \multicolumn{3}{|c|}{\textbf{same segment}}  & \multicolumn{3}{|c|}{\textbf{overall}} \\\cline{2-10}
   & \textbf{Precision}  & \textbf{Recall} & \textbf{F1} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1}\\ \hline    
 \multicolumn{10}{|c|}{lexical features only} \\ \hline   
 NB & 0.594 & 0.662 & 0.626 & 0.988 & 0.985 & 0.987 & 0.975 & 0.974 & 0.975  \\ \hline
 SVM & 0.721 & \textbf{0.705} & 0.713 & \textbf{0.990} & 0.991 & 0.990  & 0.981 & 0.981 & 0.981 \\ \hline
 KNN & \textbf{0.808} & 0.663 & \textbf{0.728} & 0.989 & \textbf{0.995} & \textbf{0.992} & \textbf{0.983} & \textbf{0.984} & \textbf{0.983} \\ \hline
 \multicolumn{10}{|c|}{lexical and puctuation features} \\ \hline
 LSTM & \textbf{0.810} & 0.720 & \textbf{0.762}  & \textbf{0.994} & \textbf{0.994} & \textbf{0.994} & \textbf{0.988} & \textbf{0.985} & \textbf{0.987} \\ \hline
 GRU & 0.783 & \textbf{0.739} & 0.760 & \textbf{0.994} & \textbf{0.994} & \textbf{0.994} & \textbf{0.988} & \textbf{0.985} & \textbf{0.987} \\ \hline
  \end{tabular}
\end{table}

As follows from Table~\ref{tab:result_base}, KNN outperforms all other models in terms of precision and F1-measure achieving 0.808 precision with 0.728 F1-measure for new segment detection. KNN also shows superior performance in all performance metrics for ``same segment'' class and weighted average over ``new segment'' and ``same segment'' classes. NB demonstrates the lowest performance among all models in terms of all performance metrics. On the other hand, SVM has the highest recall of 0.705 when only lexical features are used to identify ``new segment''. Results indicate that performance of classifiers is remarkably higher for ``same segment'' class compared to ``new segment'' class, which is expected since 96.71\% instances belong to ``same segment'' categories. For example, KNN achieves 22.40\%, 50.07\% and 36.26\% higher precision, recall and F1-measure, respectively, in ``same segment'' identification compared to ``new segment'' detection. When lexical features are used in combination with punctuation mark, LSTM demonstrates the best performance in terms of precision and F1-measure for ``new segment'' identification while GRU shows 2.64\% higher recall than LSTM. GRU and LSTM show exactly same performance in terms of all performance metrics for ``same segement'' and overall classifications. \\

@@

 NB & 0.573 & 0.680 & 0.622 & 0.573 & 0.677 & 0.621 & 0.593 & 0.662 & 0.626 \\ \hline
 SVM & 0.831 & 0.783 & 0.806 & \textbf{0.877} & \textbf{0.815} & \textbf{0.845} & 0.867 & 0.804 & 0.834  \\ \hline
 KNN & 0.808 & 0.741 & 0.773 & 0.814 & 0.732 & 0.771 & 0.814 & 0.740 & 0.775 \\ \hline
  \end{tabular}
\end{table}

Table~\ref{tab:result_boundary} summarizes the performance of NB, SVM and KNN models for detecting ``new segment'' class. We observed that classification performance significantly improved for ``new segment'' class when lexical features are used in combination with punctuation and model-based topic features. Precision increases by 15.26\% and 0\%, 21.64\% and 0.74\% and 20.25\% and 0.74\% for SVM and KNN; recall increases by 11.06\% and 17.06\%, 15.60\% and 15.64\% and 14.04\% and 16.90\% for SVM and KNN; and F1-measure increases by 13.04\% and 6.18\%, 18.51\% and 5.91\% and 16.97\% and 6.46\% for SVM and KNN when L-LDA, DL-LDA and LCA-based topic features are used in conjunction combination with lexical and punctuation features, respectively. On the other hand, NB has very negligible impact of punctuation and model-based topic features. In contrast to results in Table~\ref{tab:result_base}, SVM demonstrates the best performance among all models achieving 0.877, 0.815 and 0.845 for precision, recall and F1-measure, respectively, when DL-LDA-based topic features are used in conjunction with lexical and punctuation features. NB and KNN provide their best performance when LCA-based topic features are used, achieving 0.626 and 0.775 F1-measure, respectively.\\

@@

 NB & 0.989 & 0.983 & 0.986 & 0.989 & 0.983 & 0.986 & 0.988 & 0.985 & 0.987 \\ \hline
 SVM & 0.993 & 0.995 & 0.994 & \textbf{0.994} & \textbf{0.996} & \textbf{0.995} & 0.993 & \textbf{0.996} & \textbf{0.995} \\ \hline
 KNN & 0.991 & 0.994 & 0.993 & 0.991 & 0.994 & 0.993 & 0.991 & 0.994 & 0.993 \\ \hline
  \end{tabular}
\end{table}

Similar to baseline performance, classification performance is remarkably higher for ``same segment'' class compared to ``new segment'' class (Table~\ref{tab:result_not_boundary}). The impact of punctuation and topic model-based features is also consistent with ``new segment'' classification while improvements are not as significant as ``new segment'' class. We observed that precision increases by 0.3\% and 0.2\%, 0.4\% and 0.2\% and 0.3\% and 0.2\% for SVM and KNN; recall increases by 0.4\% and -0.1\%, 0.5\% and -0.1\% and 0.5\% and -0.1\% for SVM and KNN; and F1-measure increases by 0.4\% and 0.1\%, 0.5\% and 0.1\% and 0.5\% and 0.1\% for SVM and KNN when L-LDA, DL-LDA and LCA-based topic features are used in conjunction combination with lexical and punctuation features, respectively. Similar to results in Table \ref{tab:result_boundary}, NB and KNN demonstrate the lowest and moderate performance on the task of identifying ``same segment'', respectively. \\

@@

 NB & 0.975 & 0.973 & 0.974 & 0.975 & 0.973 & 0.974 & 0.975 & 0.974 & 0.975 \\ \hline
 SVM & 0.987 & 0.988 & 0.987 & \textbf{0.990} & \textbf{0.990} & \textbf{0.990} & 0.989 & 0.989 & 0.989 \\ \hline
 KNN & 0.985 & 0.986 & 0.985 & 0.985 & 0.986 & 0.985 & 0.985 & 0.986 & 0.986 \\ \hline
  \end{tabular}
\end{table}

Table~\ref{tab:result_weighted_avg} summarizes the weighted average results of the models for segmentation of e-Coaching emails. Overall, SVM has the best performance across all topic model-based features in terms of all performance metrics, achieves 0.990 precision, recall and F1-measure when DL-LDA-based topic features are used. Similar to results in Tables~\ref{tab:result_base}, \ref{tab:result_boundary} and \ref{tab:result_not_boundary}, NB demonstrates the lowest performance among all methods while best performance achieves with LCA-based topic features compared to other model-based topic features. Influence of the additional features is also consistent with the results in Tables~\ref{tab:result_boundary} and~\ref{tab:result_not_boundary}. Precision increases by 0.9\% and 0.2\%; recall increases by 0.9\% and 0.2\%; and F1-measure increases by 0.9\% and 0.3\% for SVM and KNN methods, respectively, when lexical feature is used in combination with punctuation and best model-based topic features.\\

\section*{Discussion}
This study is the first effort to evaluate the automatic segmentation of e-Coaching emails. Experimental results indicate that SVM is the best model among all machine learning methods considered for this study. SVM achieved 0.990 F1-measure in overall, 0.845 and 0.995 F1-measures for detecting ``new segment'' and ``same segment'', respectively. The robust performance of SVM provides the evidence that machine learning models are capable to learn conceptual information from clinical exchange. It also indicates that topic features are more important than lexical features, even when deep learning methods are employed. Although the domain of this study was intentionally quite small, we believe that our study is not limited to the e-Coaching domain and it can be successfully applied to other domains, in which discourse segmentation is a preliminary step for annotation.

@@

In this paper, results are reported by each class to avoid confusion regarding overall model performance due to severe class imbalance. In addition, standard metrics: precision, recall and F1-measure were used to eliminate doubt about the model performance since accuracy is misleading for imbalanced datasets. AUC is also utilized due to its effectiveness in measuring the quality of binary classifiers for imbalanced datasets\cite{hu2015kernelized}, which are demonstrated by the ROC curves in Figure~\ref{fig:roc-curves}. Although NB provides lowest classification results, it shows moderate AUC values, achieving 0.978 AUC when only lexical features are used and 0.977 AUC when a combination of lexical, punctuation and topic features are used. On the other hand, KNN and SVM exhibit 0.972 and 0.986 AUCs when only lexical features are used; and 0.966 and 0.980 AUCs when a combination of lexical, punctuation and topic model-based features are used. Finally, LSTM demonstrates the highest AUC value among all machine learning models, achieving 0.996 AUC, while GRU achieves only 0.2\% lower AUC than LSTM.
